Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-04-18 manuscript/manuscript.tex
Improved exponential density in manuscript

It's easy to see that the delta ratio indirectly controls how many 
times the tesseroids will be subdivided based on the variation of the 
density function.
Therefore it determines the accuracy involving the numerical error that 
the density variation introduces and also the computation time.
This raises the need to determine a maximum value of $\delta$ that 
ensures an acceptable accuracy while minimising the computation time. 
We will perform this determination with the same methodology as the one 
applied to the $D$ ratio: compare the numerical model with the 
analytical solutions for a spherical shell.

@@

A more complex density variation than a linear one could be an 
exponential function like:

\noindent where $A$, $\Delta h$ and $b$ are constants.
The gravitational potential field generated by a spherical shell with 
this kind of density function on any external point at a distance $r$ 
from its centre can be calculated as follows:

This expression will allow us  to perform comparisons between the 
analytical solutions and the numerical approximation, similar as the 
ones made for the linear density case.
Although in the exponential density scenario the problem is more 
complex due to the fact that the density-based discretization must be 
used together with the modified adaptive discretization algorithm.
This means that both the distance-size ratio $D$ and the delta ratio 
$\delta$ must be simultaneously determined.

From every $D$ and $\delta$ combination we want to find the one that 
produces a numerical error lower than the 0.1\% threshold while keeping 
an effective computation, i.e. the ($D$, $\delta)$ pair that produces 
the minimum number of discretizations and ensures an acceptable 
numerical precision.
We can propose an heuristic method to perform this task: compute the 
numerical error for every ($D$, $\delta$) points belonging to a grid on 
the $D$, $\delta$ space and specify the pair that combines an 
acceptable numerical error with an efficient computation.

@@

Because the algorithm is intended to work with default $D$ and $\delta$ 
values that ensures these two conditions for every density function, 
the default value for $D$ cannot be lower than the one determined for 
the linear function.
That's why we can only choose the ($D$, $\delta$) points on which
$D \geq D_\text{linear}$.
In this way we can divide the $D$, $\delta$ space in two subregions: 
one with $D < D_\text{linear}$ and other with $D \ge D_\text{linear}$.
We can ignore points belonging to the former one (region A on Figure 
\ref{fig:D-delta-grid-search}) and focus on the latter (regions B and C 
on Figure \ref{fig:D-delta-grid-search}).

From the $D \ge D_\text{linear}$ subregion we can explore the numerical 
error of points with $D = D_\text{linear}$ and different values of 
$\delta$.
We will call $\delta_p$ as the maximum $\delta$ at which the numerical 
error is lower than the 0.1\% threshold while keeping
$D = D_ \text{linear}$.
Greater $\delta$ values than $\delta_p$ will produce computations 
with high error, while lower $\delta$ values than $\delta_p$ will 
produce more discretizations leading to non-efficient computation.

With this definition in mind we can subdivide the subregion in two: one 
with points with $\delta \le \delta_p$ (region B on Figure 
\ref{fig:D-delta-grid-search}) and other with points with $\delta > 
\delta_p$ (region C on Figure \ref{fig:D-delta-grid-search}).

It worth mention that a greater $D$ or a lower $\delta$ 
increases the number of discretizations of the tesseroid.
Although a greater $D$ may divide it in the latitude, longitude 
and/or radial direction; a lower $\delta$ will only discretize the 
original tesseroid in the radial direction.
This means that a decrease in $\delta$ will produce a more efficient 
computation than an increase in $D$.
\todo[inline]{Should we keep this paragraph? The same is explained bellow.}

@@

Taking this conclusion into account we have determined the $\delta$ 
value needed to ensure an acceptable accuracy through a comparison 
between the numerical approximation and the analytical solution for a 
spherical shell with an exponential density.
In this specific case we proposed an exponential density function  that 
assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's 
outer and inner radii, respectively. It can be defined as follows:

@@

\noindent where $R$ is the mean Earth radius and $b$ is a constant 
that determines the variability of the function: a low value of $b$ 
increases the maximum slope of the density function.

In order to perform this comparison we must compute the analytical 
solution of the gravitational fields generated by this spherical shell 
and define a tesseroid model to numerically approximate this fields.

Firstly, the analytical solution can be computed as the sum of the 
fields generated by a spherical shell with constant density $\rho(r') = 
C$ and the one determined in equation \ref{eq:shell-pot-exp}.

Secondly, we proposed a thin and a thick spherical shells with 1km and 
35km of thickness, respectively, modelled by a mesh of $30^\circ \times 
30^\circ$ tesseroids.
We have assigned the exponential density function proposed on equation 
\ref{eq:density-exp} to each spherical shell model, choosing few values 
of the constant $b$ in order to test the numerical approximation for 
different density variations.

>>>>

It's easy to see that the delta ratio indirectly controls how many 
times the tesseroids will be subdivided based on the variation of the 
density function.
Therefore it determines the accuracy involving the numerical error that 
the density variation introduces and also the computation time: a lower 
value of $\delta$ will produce more discretizations leading to an 
increase in the accuracy of the approximation but also in the 
computation time.
This raises the need to determine a maximum value of $\delta$ that 
ensures an acceptable accuracy while minimising the computation time. 
We will perform this determination with the same methodology as the one 
applied to the $D$ ratio: compare the numerical model with the 
analytical solutions for a spherical shell.

@@

Lets perform a similar comparison between the numerical model and 
the analytical solution but now for a spherical shell with exponential 
density.
On this scenario the problem is more complex due to the fact that the 
density-based discretization will be certainly applied together with 
the modified adaptive discretization algorithm.
This means that both the distance-size ratio $D$ and the delta ratio 
$\delta$ must be simultaneously determined.

From every combination of the $D$ and $\delta$ ratios we want to find 
the one that produces a numerical error lower than the 0.1\% threshold 
while keeping an effective computation, i.e. the ($D$, $\delta)$ pair 
that produces the minimum number of discretizations and ensures an 
acceptable numerical precision.
We can propose an heuristic method to perform this task: compute the 
numerical error for every ($D$, $\delta$) points belonging to a grid on 
the $D$, $\delta$ space and choose the pair that combines an 
acceptable numerical error with an efficient computation.

@@

Because the algorithm is intended to work with default $D$ and $\delta$ 
values that ensures these two conditions for every density function, 
the default value for $D$ cannot be lower than the one determined for 
the linear function, which will be called $D_\text{linear}$.
That's why we can only choose the ($D$, $\delta$) points on which
$D \geq D_\text{linear}$.
Taking this fact into account we can divide the $D$, $\delta$ space in 
two subregions: one with $D < D_\text{linear}$ and other with $D \ge 
D_\text{linear}$. We can ignore points belonging to the former one 
(region A on Figure \ref{fig:D-delta-grid-search}) and focus on the 
latter (regions B and C on Figure \ref{fig:D-delta-grid-search}).

From the $D \ge D_\text{linear}$ subregion we can explore the numerical 
error of points with $D = D_\text{linear}$ and different values of 
$\delta$.
We will call $\delta_p$ to the maximum $\delta$ at which the numerical 
error is lower than the 0.1\% threshold while keeping
$D = D_ \text{linear}$.
Greater $\delta$ values than $\delta_p$ will produce computations with 
high error (red points in Figure \ref{fig:D-delta-grid-search}), while 
lower $\delta$ values than $\delta_p$ will produce more discretizations 
leading to non-efficient computation (green dots under the 
$(D_\text{linear}, \delta_p)$ point on Figure \ref{fig:D-delta-grid-search}).

With the definition of $\delta_p$ in mind we can subdivide the 
subregion in two: one with points with $\delta \le \delta_p$ (region B 
on Figure \ref{fig:D-delta-grid-search}) and other with points with 
$\delta > \delta_p$ (region C on Figure \ref{fig:D-delta-grid-search}).

It worth mention that either a greater $D$ or a lower $\delta$ 
increases the number of discretizations of the tesseroid.
Although a greater $D$ may divide it in the latitude, longitude 
and/or radial direction; a lower $\delta$ will only discretize the 
original tesseroid in the radial direction.
This means that a decrease in $\delta$ will produce a more efficient 
computation than an increase in $D$.
\todo[inline]{Should we keep this paragraph? The same is explained bellow.}

@@

In order to determine the value of $\delta_p$ we should compute the 
differences between the numerical model and the analytical solution for 
a spherical shell with an exponential density.
Thus we must obtain this analytical solution and define a tesseroid 
model to numerically approximate the gravitational fields.

Lets consider a decreasing exponential density function that assumes 
the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and 
inner radii, respectively, that can be defined as follows:

@@

\noindent $R$ is the mean Earth radius and $b$ is a constant 
that determines the variability of the function: a low value of $b$ 
increases the maximum slope of the density function.

Firstly, the analytical solution of the gravitational fields generated 
by a spherical shell with a density functions as the one in equation 
\ref{eq:density-exp} can be computed as the sum of the fields generated 
by a spherical shell with constant density $\rho(r') = C$ and another 
shell with an exponential density like:

\noindent where $A$, $\Delta h$ and $b$ are constants.
The gravitational potential field generated by a spherical shell with 
this density on any external point at a distance $r$ from its centre 
can be calculated by solving the integral in equation \ref{eq:shell-pot},
resulting as follows:

Secondly, we proposed a thin and a thick spherical shells with 1km and 
35km of thickness, respectively, modelled by a mesh of $30^\circ \times 
30^\circ$ tesseroids.
We have assigned the exponential density function proposed on equation 
\ref{eq:density-exp} to each spherical shell model, choosing some values 
of the constant $b$ in order to test the numerical approximation for 
different density variations.
